\( \tool \): JavaScript IR-based Semantics Extraction Toolchain

### 2019-08-20 icse/compiler.tex
Fixed error of code in algorithm copiler section

\section{Algorithm Compiler}\label{sec:compiler}

\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{img/algo_compiler.png}
  \caption{Overall structure of the algorithm compiler.}
  \label{fig:algo-compiler}
\end{figure}

In this section, we propose the \textit{algorithm compiler}
that compiles abstract algorithms into \( \ires \) functions
and its overall structure is described in Figure~\ref{fig:algo-compiler}.
Before compiling abstract algorithms,
the tokenizer first tokenizes each abstract algorithm into
a list of tagged tokens. Then, the token parser constructs
syntactic parse tree of the given token list.
Finally, the \( \ires \) function generator constructs the corresponding
\( \ires \) function from the given parse tree.
The parser and the generator is dependent on conversion rules
and we define the general conversion rules for abstract algorithms in
ECMAScript specifications. Like Coq, a proof assistant, we also provide
the \textit{rule generation assistant} to make easy write conversion rules.
It suggests new conversion rules based on
statistical analysis of algorithm steps that failed to be parsed
and also diagnoses root causes of parsing failures.

\subsection{Tokenizer}

Abstract algorithms in ECMAScript specifications are written in structured
natural languages in HTML files. An algorithm consists of ordered steps
and it might contain sub-steps as well. For example,
the \( \code{ToPrimitive} \) algorithm in Figure~\ref{fig:to-primitive-es}
has three steps and second step has seven sub-steps.
Moreover, the tokens of each step has its own HTML tag and each tag
has the following meaning:
\[
  \begin{array}{c|l}
    \text{HTML tags} & \text{meanings}\\\hline\hline
    \code{<var>} & \text{variables}\\\hline
    \code{<emu-grammar>} & \text{productions}\\\hline
    \code{<emu-nt>} & \text{non-terminal syntax}\\\hline
    \code{<code>} & \text{ECMAScript codes}\\\hline
    \code{<emu-const>} & \text{constant values}\\\hline
    \code{<emu-val>} & \text{values}\\\hline
    \code{<ol>} & \text{ordered sub-steps}\\\hline
    \code{<ul>} & \text{unordered sub-steps}\\\hline
    \code{<sup>} & \text{superscripts}\\\hline
    \text{otherwise} & \text{simple texts}\\\hline
  \end{array}
\]
We try to keep the tag information for each token to generate more
precise conversion rules. For example, if and only if a token has a tag
\( \code{<var>} \), it is a parameter or a local variable.
Thus, it is possible to construct a conversion rule precisely discriminates
identifiers and other components.

The tokenizer first recognizes the overall structures of steps.
Then, it divides each step into sequence of tagged tokens.
If an HTML element has a explicit tag, it is converted into a single token
with the tag. Otherwise, it is divided into multiple tokens and each token
should be a sequence of alphanumeric characters or a single
non-alphanumeric character. For example, in the \( \code{ToPrimitive} \)algorithm,
\( \textbf{\code{"default"}} \) is a single token with the tag \( \code{<code>} \)
and \( \code{@@toPrimitive} \) is divided into three text tokens
\( \code{@} \), \( \code{@} \), and \( \code{toPrimitive} \).

For the linear structures, the tokenizer flatten the structured steps into
a single list. Some conversion rules should handle multiple steps such as
the conditional statements (if-then-else). Thus, we decide to break down
structured algorithms using three special tokens;
\( \tend \) denotes the end of a single step,
\( \tin \) and \( \tout \) the start and the end of nested steps, respectively.
For example, the \( \code{ToPrimitive} \) algorithm is tokenized as follows.
\[
  \begin{array}{l}
    \code{Assert} \cdots \tend\\
    \code{If} \cdots \tin \code{If} \cdots \tend
    \cdots \code{Return} \cdots \tend \tout \tend\\
    \code{Return} \cdots \tend\\
  \end{array}
\]

\subsection{Token Parser}

The token parser parses a given list of tokens into a syntactic parse tree.
It depends on the given \textit{conversion rules} that consists of
two parts parsing rules and mapping from parsing rules into corresponding
\( \ires \) components.
A parsing rule consists of basic token matchers and other parsing rules
including itself. For example, the following parsing rule is simplified version for
sub-step \( i \) in the \( \code{ToPrimitive} \) algorithm.
\begin{lstlisting}[style=myScalastyle]
// statements
val Stmt = "Let" ~ varT ~ "be" ~ Expr ~ "." ^^ ...
// expressions
val Expr = (
  // identifiers
  varT ^^ ... |
  // return if abrupt
  "?" ~ Expr ^^ ... |
  // function calls
  textT ~ "(" ~ repsep(varT, ",") ~ ")" ^^ ... |
  // lists
  "<<" ~ repsep(Expr, ",") ~ ">>" ^^ ...
)
\end{lstlisting}

Each parsing rule is written in extended Scala parser combinators.
We modify the meaning of alternative composition operator ( \( | \) ) to collect
all longest matched results. If the parser detects that a step cannot be
parsed with the given parsing rules. It reports that a step is not possible
to be parsed under the given rules into the rule generation assistant.
Moreover, even though the parser could parse a step, it will be also reported
if it could be parsed in not a single but multiple ways.
Each string literal is a token matcher for tokens without any tags and checks
that the token has same string value with the string literal.
The \( \code{varT} \) and \( \code{textT} \) is also token matchers
for token with the tag \( \code{<var>} \) and no tags.
The \( \code{Step} \) and \( \code{Expr} \) are user-defined parsing rules
defined with our extended parser combinators.
Moreover, it supports all helpers functions defined in Scala parser combinators.
For example, the helper function \( \code{repsep(p, q)} \) generates a new
parsing rule that denotes zero or more repetitions of the parsing rule \( \code{p} \)
using another parsing rule \( \code{q} \) as separators.
Finally, the token parser with the above rules parses sub-step \( i \) in the
\( \code{ToPrimitive} \) into the following syntactic parse tree:
\includegraphics[width=0.5\textwidth]{img/token_ast.png}

\subsection{\( \ires \) Function Generator}

We define \( \ires \) language to give semantics into syntactic parse tree of
abstract algorithm steps. It has the following language features:

\begin{itemize}
\item \textbf{Dynamic Typing} We decide to use dynamic typing for \( \ires \)
because each variables in abstract algorithms are not statically typed.
Thus, variables do not have its own static types while each value of \( \ires \)
has its dynamic type.

\item \textbf{Imperative Style} \( \ires \) represents algorithm steps
as statements. Each statement changes the current state
that consists of environments and a heap.

\item \textbf{Higher-order Functions with Restricted Scopes} \( \ires \) supports
lambda functions as values but it has only restricted scopes. In each function,
only global variables, parameter, and newly defined local variables are accessible.
It means that the function closure does not capture the the current environment.
We use such restricted scopes because it is enough to represent abstract algorithms.

\item \textbf{Primitive Values} \( \ires \) supports ECMAScript primitive values except
the symbols that is newly introduced after ECMAScript 6. Thus, only five ECMAScript primitives
are defined in \( \ires \); numbers, strings, booleans, \( \code{undefined} \), and \( \code{null} \).
Moreover, \( \ires \) also supports the unique \( \code{absent} \) value to represent
the absence of parameters. For example, the second parameter \( \code{PreferredType} \)
of \( \code{ToPrimitive} \) algorithm in Figure~\ref{fig:to-primitive-es} might be
not present in some function calls. Then, the parameter has the \( \code{absent} \) value
to represent its absence.

\item \textbf{Abstract Data Types} \( \ires \) only supports three abstract data types;
\( \code{Record} \) for mapping from some values into other value,
\( \code{List} \) for sequential data types,
and \( \code{Singleton} \) for singleton data types.
They are stored in corresponding addresses in the heap structure.
For example, ECMAScript environment records are represented as \( \code{Record} \) data types
from String values for identifier names into addresses.
The addresses point to other \( \code{Record} \) data types that represent ECMAScript bindings.
We treat a ECMAScript constants such as \( \code{empty} \) and primitive symbols as singleton
data types in \( \ires \).
\end{itemize}

To generate the corresponding \( \ires \) function from syntactic parse tree,
we describe the mapping from each parsing rule into the corresponding \( \ires \) components.
For example, we describe the following mapping for the previous example parsing rules:
\begin{lstlisting}[style=myScalastyle]
// statements
val Stmt = ... ^^ { case _ ~ x ~ _ ~ y => ESLet(x, y) }
// expressions
val Expr = (
  // identifiers
  ... ^^ { x => x } |
  // return if abrupt
  ... ^^ { _ ~ e => ESReturnIfAbrupt(e) } |
  // function calls
  ... ^^ { case x ~ _ ~ y ~ _ => ESCall(x, y) } |
  // lists
  ... ^^ { case _ ~ x ~ _ => ESList(e) }
)
\end{lstlisting}
Finally, sub-step \( i \) in the \( \code{ToPrimitive} \) is converted
into the following instruction:
\begin{lstlisting}[style=ires]
let result = ? (Call exoticToPrim input (new [hint]))
\end{lstlisting}

\subsection{Rule Generation Assistant}

\inred{
To make easy write conversion rules, we support \textit{rule generation assistant}.
It consists of two helper functions; the \textit{rule recommender} and
the \textit{fault localizer}. The

It makes easy not only construct conversion rules from the scratch
but also updates existing conversion rules to handle specification updates.
We explain each helper function and how to utilize it to construct conversion rules.

\subsubsection{Statistics Monitor}

TODO

\subsubsection{Fault localizer}

TODO
}

>>>>

\section{Algorithm Compiler}\label{sec:compiler}
\begin{figure}
  \centering
  \includegraphics[width=0.5\textwidth]{img/algo_compiler.png}
  \caption{Overall structure of {\sf Algorithm Compiler}}
  \label{fig:algo-compiler}
\end{figure}

In this section, we explain {\sf Algorithm Compiler} that compiles
abstract algorithms to \( \ires \) functions as illustrated in
Figure~\ref{fig:algo-compiler}.

\subsection{Tokenizer}
Before compiling abstract algorithms, {\sf Tokenizer} first tokenizes
each abstract algorithm into a list of tagged tokens.  An algorithm
consists of ordered steps and it may contain sub-steps as well.
For example, the \textbf{\small ToPrimitive} algorithm in
Figure~\ref{fig:to-primitive}(a) has three steps and the second step
has seven sub-steps.  Moreover, the tokens of each step have their own
HTML tags and each tag has a meaning.  For instance, \code{<var>}
denotes variables, \code{<emu-grammar>} denotes productions, and
\code{<code>} denotes ECMAScript code.
% \[
%   \begin{array}{c|l}
%     \text{HTML tags} & \text{meanings}\\\hline\hline
%     \code{<var>} & \text{variables}\\\hline
%     \code{<emu-grammar>} & \text{productions}\\\hline
%     \code{<emu-nt>} & \text{non-terminal syntax}\\\hline
%     \code{<code>} & \text{ECMAScript codes}\\\hline
%     \code{<emu-const>} & \text{constant values}\\\hline
%     \code{<emu-val>} & \text{values}\\\hline
%     \code{<ol>} & \text{ordered sub-steps}\\\hline
%     \code{<ul>} & \text{unordered sub-steps}\\\hline
%     \code{<sup>} & \text{superscripts}\\\hline
%     \text{otherwise} & \text{simple texts}\\\hline
%   \end{array}
% \]
We keep such HTML tag information for each token to generate more precise
{\sf Compile Rules}.  For example, because if and only if a token has
a tag \( \code{<var>} \), it is a parameter or a local variable, we can
construct {\sf Compile Rules} that precisely distinguish identifiers
from other components.

The {\sf Tokenizer} module recognizes the structure of each step,
and divides the step into a sequence of tagged tokens.
If an HTML element has an explicit tag, it is converted to a single token
with the tag.  Otherwise, it is divided into multiple tokens and each token
becomes a sequence of alphanumeric characters or a single non-alphanumeric
character.  For example, in the \textbf{\small ToPrimitive} algorithm,
\( \textbf{\code{"default"}} \) is a single token with the HTML tag \( \code{<code>} \)
and \( \code{@@toPrimitive} \) is divided into three text tokens
\( \code{@} \), \( \code{@} \), and \( \code{toPrimitive} \).

Moreover, {\sf Tokenizer} flattens a structured step to a single token
list to handle multi-step statements easily.  Some statements in
abstract algorithms consist of multiple steps.  For example, the
\( \code{if}\!-\!\code{then}\!-\!\code{else} \) statement often consists of
two steps: one for the \( \code{then} \)-branch and another
for the \( \code{else} \)-branch.  To treat them as a linear structure,
we introduce three special tokens to break down structured algorithms:
\( \tend \) denotes the end of a single step, and \( \tin \) and
\( \tout \) denote the start and the end of nested steps, respectively.
For example, the \textbf{\small ToPrimitive} algorithm is tokenized as follows.
\[
  \begin{array}{l}
    \code{Assert} \cdots \tend\\
    \code{If} \cdots \tin \code{If} \cdots \tend
    \cdots \code{Return} \cdots \tend \tout \tend\\
    \code{Return} \cdots \tend\\
  \end{array}
\]

After tokenizing abstract algorithms, {\sf Algorithm Compiler}
compiles token lists into \( \ires \) functions using
{\sf Token List Parser} and {\sf Token AST Converter}.
They depend on {\sf Compile Rules} and each compile rule
consists of a \textit{parsing rule} and a \textit{conversion rule}:
\begin{lstlisting}[style=myScalastyle]
val CompileRule = ParsingRule ^^ ConversionRule
\end{lstlisting}
For each compile rule, its parsing rule describes how to parse a given
token list into a structured token AST, and its conversion rule describes
how to convert the given token AST structure into an \( \ires \) component.
Now, we explain the token list parser and token AST converter with
parsing rules and conversion rules, respectively.

\subsection{Token List Parser}
Parsing rules determine the token list parser, and each parsing rule
is written in extended Scala parser combinators.  We modified the
meaning of the alternative composition operator ( \( | \) ) to collect
all the longest matched results.  If a parser detects a step that
cannot be parsed with given parsing rules, it reports the step into
{\sf Rule Assistant}.  Even when a parser parses a step successfully,
it also reports the step if it can parse the step in multiple ways.
We use Packrat parsing~\cite{packrat} for parsing rules, which
support left recursion.

We provide two kinds of basic token parsers: tag-based parsers and
content-based parsers.  A tag-based parser just checks whether the next
token has a given tag.  For example, the tag-based parser \( \code{varT} \)
checks whether the next token has the tag \( \code{<var>} \).  A
content-based parser checks whether the next token is a text token and
its content passes a given condition.  For example, the string literal
\( \code{"Let"} \) denotes a content-based parser that checks whether
the next token is a text token with the content \( \code{Let} \).
We also provide two content-based parsers \( \code{word} \) and
\( \code{number} \) that check whether the content of the next token
consists of only alphabets or numbers, respectively.  In parsing
rules, all the helper functions defined in Scala parser combinators
are available.  For instance, the helper function \( \code{repsep(p, q)} \)
generates a new parsing rule that denotes zero or more repetition of the
parsing rule \( \code{p} \) using another parsing rule \( \code{q} \) as
a separator.

Consider the following example parsing rule, which is a simplified
version for the step 2-e-i of the \textbf{ToPrimitive} algorithm.
\begin{lstlisting}[style=myScalastyle]
// statements
val Stmt = "Let" ~ varT ~ "be" ~ Expr ~ "." ^^ ...
val Expr = (          // expressions
  varT ^^ ... |       // identifiers
  "?" ~ Expr ^^ ... | // return if abrupt
                      // function calls
  word ~ "(" ~ repsep(Expr, ",") ~ ")" ^^ ... |
  "<<" ~ repsep(varT, ",") ~ ">>" ^^ ...  // lists
)
\end{lstlisting}
The \( \code{Stmt} \) compile rule describes how to compile statements
with one parsing rule and its conversion rule is omitted for now for
brevity.  The \( \code{Expr} \) compile rule describes how to compile
expressions with four parsing rules.  A token parser with the above
rules parses the step 2-e-i of \textbf{ToPrimitive} to the following
token AST:
\begin{center}
\includegraphics[width=0.5\textwidth]{img/token_ast.png}
\end{center}

\subsection{Token AST Converter}
Conversion rules describe how to generate an \( \ires \) function for a
given token AST.  Each conversion rule is defined with its
corresponding parsing rule.  For basic token parsers, their conversion
rules always return the string values of the contents in parsed tokens.
For example, the following conversion rules are the omitted parts
in the previous example for the step 2-e-i of \textbf{ToPrimitive}:
\begin{lstlisting}[style=myScalastyle]
// statements
val Stmt = ... ^^ { case _ ~ x ~ _ ~ y => ILet(x, y) }
val Expr = (               // expressions
  ... ^^ { x => EId(x) } | // identifiers
                           // return if abrupt
  ... ^^ { _ ~ e => EReturnIfAbrupt(e) } |
                           // function calls
  ... ^^ { case x ~ _ ~ y ~ _ => ECall(x, y) } |
  ... ^^ { case _ ~ x ~ _ => EList(e) } // lists
)
\end{lstlisting}
The conversion rule of the compile rule \( \code{Stmt} \) uses only
the second and fourth sub trees and converts them to \( \ires \)
components.  The second sub-tree is converted to a string value
because it was constructed via the basic token parser \( \code{varT} \).
The fourth sub-tree is converted to an \( \ires \) expression
because the conversion rules of the \( \code{Expr} \) compile rule
construct \( \ires \) expressions.  Then, the conversion rule of
\( \code{Stmt} \) constructs the lexical binding instruction \( \code{ILet} \)
with the string value and the \( \ires \) expression.  In this way,
the step 2-e-i of \textbf{ToPrimitive} is converted to the following
instruction:
\begin{lstlisting}[style=ires]
let result = ? (Call exoticToPrim input (new [hint]))
\end{lstlisting}

We define \( \ires \) to represent abstract algorithms as its
functions with the following design choices:

\begin{itemize}[leftmargin=0.5cm]
\item \textbf{Dynamic typing:} Because each variable in abstract
algorithms is not statically typed, variables do not have their own
static types while each value of \( \ires \) has its dynamic type.

\item \textbf{Imperative style:} \( \ires \) represents algorithm steps
as imperative instructions in the sense that each instruction changes
the current state consisting of an environment and a heap.

\item \textbf{Higher-order functions with restricted scopes:}
In each function of \( \ires \), only global variables, parameters,
and its local variables are available, which means that a function
closure does not capture its current environment.  We use such
restricted scopes because they are enough to represent abstract
algorithms.

\item \textbf{Primitive values:} \( \ires \) supports ECMAScript
primitive values except ``symbols'' because symbols can be represented
as singleton objects.  Also, \( \ires \) provides the unique \(
\code{absent} \) value to represent the absence of parameters.  For
example, when the optional second parameter \( \code{PreferredType} \)
of \textbf{ToPrimitive} in Figure~\ref{fig:to-primitive}(a) is absent,
the parameter has the \( \code{absent} \) value.

\item \textbf{Abstract data types:} \( \ires \) supports only three
abstract data types: \( \code{Record} \) for mappings from some values
to other values, \( \code{List} \) for sequential data, and
\( \code{Singleton} \) for singleton data.  For example, ECMAScript
environment records are represented as \( \code{Record} \)
with string values as keys and addresses as values.  Each key
represents an identifier name and its corresponding address represents
an ECMAScript object.  We also represent constants in the ECMAScript
specification such as \( \code{empty} \) and \( \code{normal} \) as
\( \code{Singleton} \) data types in \( \ires \).
\end{itemize}

\subsection{Rule Assistant}
\begin{algorithm}[t]
  \caption{Fault Localization}\label{alg:fault}
  \begin{algorithmic}[1]
    \Function{FaultLocalize}{$T,n$}\Comment{$T$: failed step, $n$: size}
    \State $S = \varnothing$
    \For{$i = 0$ to $n - 1$}
    \For{$j = i$ to $n - 1$}
    \State $T'$ = a copy of $T$
    \State Substitute $T'[i..j]$ with $\tstar$
    \If{Parse($T'$) contains $\code{Dummy(}x \code{)}$ for some $x$}
    \State Insert $(x, i, j)$ into $S$
    \EndIf
    \EndFor
    \EndFor
    \While{$s = (x, i, j) \in S$}
    \If{$\exists (y, p, q) \in S$ s.t. $i \leq p \wedge q \leq j$}
    \State Remove $s$ from $S$
    \EndIf
    \EndWhile
    \State \Return $S$
    \EndFunction
  \end{algorithmic}
\end{algorithm}

To make it easy to write compile rules, we support a \textit{rule assistant},
which takes not-yet-convertible steps dubbed ``failed steps'' as arguments.
For each failed step, the rule assistant diagnoses a root cause of the
parsing failure.  The basic idea is to check whether we can parse the
failed input with the assumption that a specific sub-sequence of the
failed step is successfully parsed.  We define an approach to find
such sub-sequences of failed steps as \textit{fault localization}.
This approach was actually helpful for us when we update existing
compile rules for new versions of the specification.

The fault localization algorithm is shown in Algorithm~\ref{alg:fault}.
For a given failed step $T$ and the size of a failed input $n$, it
attempts to find a set of the most specific suggestions.  It
substitutes a specific sub-sequence \( T[i..j] \) of \( T \) with the
special token \( \tstar \), which represents some sub-sequence of a
failed step.  It collects a set of possible suggestions, and then
removes more general ones than existing suggestions.  To support this
process, we added a parsing rule  \( \code{starT} \) to each compile
rule as follows:
\begin{lstlisting}[style=myScalastyle]
// statements
val Stmt = ... | starT ^^ { case x => Dummy("Stmt") }
// expressions
val Expr = ... | starT ^^ { case x => Dummy("Expr") }
\end{lstlisting}
where \( \code{starT} \) is a special basic token parser that just
checks whether a given token is the special token \( \tstar \). 
Its corresponding conversion rule is a function that returns a dummy
\( \ires \) component \( \code{Dummy} \) with a string value that
represents the name of its compile rule.

For example, let us assume that the \( \code{Expr} \) compile rule
does not have the alternative for lists:
\begin{lstlisting}[style=myScalastyle]
  "<<" ~ repsep(varT, ",") ~ ">>" ^^
       { case _ ~ x ~ _ => EList(e) } // lists
\end{lstlisting}
Then, the step 2-e-i of \textbf{ToPrimitive} becomes a failed step.
When we use the fault localization algorithm with the failed step and
its size 15, it first finds one suggestion for statements and four
suggestions for expressions as follows:
\[
\footnotesize
\! \! \! \!
  \begin{array}{c}
    (\scode{"Stmt"}, 0, 14), \;
    (\scode{"Expr"}, 3, 13), \;
    (\scode{"Expr"}, 4, 13), \;
    (\scode{"Expr"}, 6, 12), \;
    (\scode{"Expr"}, 10, 12)
  \end{array}
\]
In the \textbf{\small while} loop on lines 9--11, because the last one
is the most specific sub-sequence, it removes the other four
suggestions. Finally, the rule assistant suggests to define another
expression compile rule for the token list \hint{}.
